### 2020-05-24 livelits-paper.tex
separated everything into separate files

\input{intro}

\input{overview}

%% \section{Defining Palettes}

\section{Overview Examples: Defining Palettes}

>>>>

\input{intro}
\input{case-studies}
\input{livelit-definitions}
\input{livelit-calculus}
\input{extensions}
\input{implementation}
\input{related-work}
\input{discussion}

\clearpage
\appendix
\section{Misc. Material from Prior Draft}

To address these limitations, we introduce live literals, or livelits. Figure~\ref{fig:cutoffs} shows a mockup of the definition and application of a livelit named \li{$grade_cutoffs} for adjusting grade cutoffs, represented as values of  record type \li{grade_cutoffs}. Like textual literal forms (e.g. list literals), livelits are alternative representations of expressions of the associated type \cite{DBLP:journals/pacmpl/OmarA18}. % From the perspective of the remainder of the program, \li{cutoffs} is a value of type \li{grade_cutoffs} like any other.
We are implementing livelits in Hazel (\url{hazel.org}), a live functional programming environment with support for typed holes \cite{popl-paper}. We plan to perform a live demo.

The definition of
the livelit, outlined in Figure~\ref{fig:cutoffs}, follows the Elm architecture,
i.e. there are types representing the abstract model and the messages that the 
GUI generates. Livelits are \textbf{persistent} rather than ephemeral, i.e. the model is recorded in the underlying syntax tree. The \li{view} function generates the GUI, implemented using HTML, on demand (so the view is not persisted). Another function, \li{to_exp}, is responsible for generating the underlying expression, called the \emph{expansion}, from the \li{model}. 
While other projectional editors, e.g. those generated by Citrus \cite{DBLP:conf/uist/KoM05}, also support persistent GUIs in code, they are not user-extensible.

The GUI can itself contain typed holes, represented by the \li{HtmlHole} constructor. In this case, there is a single hole
for entering an expression of type \li{list(float)}, i.e. the list of weighted
averages. In this example, the user has filled this hole with a variable,
\li{weighted_averages} (the definition of which precedes the content of this figure and is not shown). In other words,
livelits support \textbf{open expressions} and therefore interact cleanly with 
standard abstraction mechanisms, i.e. they can appear under binders. We follow the reasoning principles for literals with spliced sub-expressions established  by \citet{DBLP:journals/pacmpl/OmarA18}.

The main complication when dealing with open expressions relates to how live feedback
is to be generated. Given just the symbolic expression in the hole, it would be 
impossible to plot (as orange dots) the actual data from the list that the variable refers to. To resolve this issue, the system evaluates the program 
as if it the livelits were empty holes, 
relying on the support for evaluating incomplete programs described recently 
by \citet{DBLP:journals/pacmpl/OmarVCH19}. The result of evaluation is an expression containing
hole closures, i.e. holes equipped with environments. 
%Hazel allows the user to interactively select which closure
%interests them if multiple closures for a given hole appear in the result (e.g. if a livelit appears inside a function called multiple times). 
The livelit 
can then evaluate the expression in the hole against the closure selected by the user (not shown) using \li{Env.run}.

%%%%%

\todo{this is the WIP one from last year}{}
Programs are most often represented as text.
%
This representation format provides expert programmers access to a variety of
text editors and text edit actions---to affect fine-grained control over
programming decisions---and a large ecosystem of other text-based
utilities---for example, for line-based file differencing of source code
versions.
%
On the other hand, the flexibility and low-level nature of text comes at some
costs.
%
For experts, some tedious and manual edits could lead to
inefficiency---especially because many routine code editing operations do not
require the full flexibility of text~\citep{XXX}.
%
For novices, the large class of syntax errors that stem from text-edits presents
a steep learning curve.

Motivated to overcome the disadvantages of represent programs as text, a variety
of alternative code editing interfaces have been investigated.
%
At the opposite end of text are \emph{visual programming languages}, which often
completely represent a program with graphical elements~\citep{XXX}.
%
To forgo text completely, such approaches often target domain-specific
languages, such as dataflow programing~\citep{XXX} and pedagogical languages
that are, by design, restricted to relatively few building blocks.
%
Other \emph{structure}, or \emph{projectional} editors, still use a significant
amount of text to render programs, but forgo text-edits in favor of
\emph{structure edit actions} which transform the program (represented as a tree
or some other structure), sidestepping the danger of invalid intermediate states
of concrete syntax.

On the spectrum somewhere between fully text- and fully structure-based are
``hybrid'' editors, which augment text with additional ways to visualize and
manipulate the structure of the program.
%
Victor Scrubbing, APX~\citep{APX}, Sketch-n-Sketch~\citep{sns-pldi}, and Carbide
IDE~\citep{XXX}, among others, allow numeric values to be ``scrubbed'' by
directly manipulating sliders rather than text-editing numeric literals.
%
Barista~\cite{Barista} is a hybrid Java editor where custom \emph{structure
view} GUIs provide alternate representations of expressions instead of text.
%
For example, an arithmetic expression may be rendered with mathematical symbols,
a method may be accompanied by interactive documentation with input-output
examples, and structures may be selectively collapsed, expanded, or zoomed.
%
Graphite~\citep{Graphite} allows custom GUIs---called \emph{palettes}---to help
the programmer fill missing expressions---``holes''---in the program.
%
For example, a color palette can provide visual previews of different candidate
color values, and a regular expression palette can show input-output examples
for different candidate regular expressions.

The GUI representations and interactions enabled by the above hybrid editors are
useful, but several limitations likely preclude wider utility:
%
the types of expressions that benefit from alternative GUI editors are limited
to
%
privileged types that have baked-in interfaces~\citep{XXX}
%
or just of base type~\citep{XXX};
%
the GUI view is ephemeral, in that it disappears once it has been used to
generate an expression~\citep{XXX}; or
%
expressions generated by the GUI are not deeply integrated with the static type
system and interpreter of the language~\citep{XXX,XXX,XXX}.

%% refactoring
%% 
%% DNDRefactoring~\citep{DNDRefactoring}
%% Deuce~\citep{sns-deuce}

\parahead{Persistent, Composable, and Live GUIs for Filling Holes}

compared to the ``simple'' palettes

extend palettes with: persistence, composition, and live feedback

macro systems allow alternative syntactic (i.e. string) representations, to
expand into underlying expressions

palatable

by analogy to these ``literal'' macro systems, palettes are ``graphical
macros'': through interaction with the user, the GUI generates the underlying
expression.

TLMs~\cite{TLMs}

\parahead{Contributions and Paper Outline}

design for palettes. specifically, within the Hazelnut Live framework,
which is a system to address the gap problem that arises in traditional editor
frontends.

demonstration of the expressiveness of the approach through a series of
examples, many of which are drawn from the user study mentioned before.

a prototype implementation of palettes within Hazel, which currently
supports several examples using a core calculus with minimal features. the
implementation provides a clear path for scaling up to larger, full-featured
syntactic programming conveniences, as well as further UI design in future work.

prototype UIs:

  \begin{enumerate}
    \item \Hazel{}: (palette macros;) structure edit actions; expression formula bar; inline layout
    \item \sns{}: (palette functions;) text editor; pop-up menu
  \end{enumerate}

hazelnut~\citep{Hazelnut,HazelnutLive}

\begin{figure*}[t!]
  \begin{center}
  \includegraphics[width=34pc]{cutoffs-mockup.png}\end{center}
  \caption{A mockup of a livelit for adjusting grade cutoffs. Livelits are persistent and can access the live environment.}
  \label{fig:cutoffs}
  \end{figure*}

%% \section{Defining Palettes}
\input{overview}
\section{Overview Examples: Defining Palettes}
